Measurement Clusterization in Bayesian D-optimal Designs in Infinite Dimensions

### 2024-07-31 latex/how.tex
before diff
Intro ready

%\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
\section{Substituting an optimal clustered design with an equally optimal non-clustered design}\label{section:how}
%
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which might allow us to avoid a
clustered design. As an example, consider an inverse problem for which
the first two eigenvalues of the pushforward prior $\fwd \prcov\fwd^*$
are $\lambda_1 = 10, \lambda_2 = 5$, with eigenvectors
$\{\ev_i\}_{i=1}^\infty$ and $\sigma^2=1$. Consider the following
$\obs, \pobs$ (see also Figure \ref{fig:clusterization}):
\begin{equation}\label{eq:two designs}
  \obs =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \pobs =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs^*\obs = \pobs^*\pobs$
and that both $\obs$ and $\pobs$ are D-optimal. However, $\obs$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\pobs$ is
not.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs$ in \eqref{eq:two designs})
    and non-clusterization (right, $\pobs$ in \eqref{eq:two designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs$ and $\pobs$ from \eqref{eq:two
      designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}

>>>>

%\section{Clusterization poses no obstruction to D-optimality}\label{section:clusterization}
\section{Substituting an optimal clustered design with an equally optimal non-clustered design}\label{section:how}
%
The necessary conditions for D-optimality of Theorem \ref{thm:char}
leave some freedom in choosing $\obs$, which might allow us to avoid a
clustered design. As an example, consider an inverse problem for which
the first two eigenvalues of the pushforward prior $\fwd \prcov\fwd^*$
are $\lambda_1 = 10, \lambda_2 = 5$, with eigenvectors
$\{\ev_i\}_{i=1}^\infty$ and $\sigma^2=1$. Consider the following
$\obs, \pobs$ (see also Fig.~\ref{fig:clusterization}):
\begin{equation}\label{eq:two_designs}
  \obs =
  \begin{pmatrix}
    \sqrt{\frac{11}{40}}\ev_1 - \sqrt{\frac{29}{40}}\ev_2 \\
    \sqrt{\frac{11}{40}}\ev_1 + \sqrt{\frac{29}{40}}\ev_2 \\
    \ev_1 \\
    \ev_2 \\
    \ev_1 \\    
  \end{pmatrix},
  \pobs =
  \begin{pmatrix}
    \sqrt{\frac{3}{8}}\ev_1 - \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{3}{8}}\ev_1 + \sqrt{\frac{5}{8}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 - \sqrt{\frac{3}{5}}\ev_2 \\
    \sqrt{\frac{2}{5}}\ev_1 + \sqrt{\frac{3}{5}}\ev_2 \\
    \ev_1 \\    
  \end{pmatrix}
\end{equation}
\noindent It is easy to verify that $\obs^*\obs = \pobs^*\pobs$
and that both $\obs$ and $\pobs$ are D-optimal. However, $\obs$
is a clustered design ($\meas_3 = \meas_5 = \ev_1$), while $\pobs$ is
not.

@@

      \addlegendentry{prior}
      \addlegendentry{$o_1o_2$}
      \addlegendentry{$o_3o_4$}
      \addlegendentry{$o_5$}
      %% \addlegendentry{$o_4$}
      %% \addlegendentry{$o_5$}
      %% \addlegendentry{$o_6$}   
    \end{axis}
  \end{tikzpicture}
  \caption{Clusterization (left, $\obs$ in \eqref{eq:two_designs})
    and non-clusterization (right, $\pobs$ in \eqref{eq:two_designs})
    in D-optimal designs. Posterior precision per eigenvector is
    plotted for $\obs$ and $\pobs$ from \eqref{eq:two
______designs}. Both designs are identical for all practical matters
    --- their posteriors are equal. In the left panel, a D-optimal
    design with clusterization is illustrated: $\meas_3 = \meas_5$. In
    the right panel, a D-optimal design without clusterization is
    illustrated. The contribution to the posterior precision due to
    $\meas_1$ and $\meas_2$ is denoted $\meas_1\meas_2$. These
    contributions are plotted together because neither $\meas_1$ nor
    $\meas_2$ are not in the direction of any eigenvector. Similarly
    for $\meas_3\meas_4$.}
  \label{fig:clusterization}
\end{figure}
